---
title: "Data Exploration"
author: "Heather Anderson"
format:
  html:
    link-external-icon: true
    link-external-newwindow: true
course: Capstone Projects in Data Science
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
theme: cyborg
---

## Dataset 1

<https://www.kaggle.com/datasets/sriharshaeedala/airline-delay?resource=download>

```{r}
AirlineDelays <- read.csv("Airline_Delay_Cause.csv")
head(AirlineDelays)
```

```{r}
dim(AirlineDelays)
```

This dataset provides information on flight arrivals and delays at U.S.
airports, sorted by airline carriers. It includes details like the
number of arriving flights, delays over 15 minutes, cancellations, and
diversions. It also shows the reasons for delays, such as airline
issues, weather, airspace factors, security, and late arrivals. You can
use this dataset to analyze how different airlines perform at various
airports and understand what causes delays in the aviation industry.

**Columns:**

-   **year:** The year of the data.

-   **month:** The month of the data.

-   **carrier:** Carrier code.

-   **carrier_name:** Carrier name.

-   **airport:** Airport code.

-   **airport_name:** Airport name.

-   **arr_flights:** Number of arriving flights.

-   **arr_del15:** Number of flights delayed by 15 minutes or more.

-   **carrier_ct:** Carrier count (delay due to the carrier).

-   **weather_ct:** Weather count (delay due to weather).

-   **nas_ct:** NAS (National Airspace System) count (delay due to the
    NAS).

-   **security_ct:** Security count (delay due to security).

-   **late_aircraft_ct:** Late aircraft count (delay due to late
    aircraft arrival).

-   **arr_cancelled:** Number of flights canceled.

-   **arr_diverted:** Number of flights diverted.

-   **arr_delay:** Total arrival delay.

-   **carrier_delay:** Delay attributed to the carrier.

-   **weather_delay:** Delay attributed to weather.

-   **nas_delay:** Delay attributed to the NAS.

-   **security_delay:** Delay attributed to security.

-   **late_aircraft_delay:** Delay attributed to late aircraft arrival.

```{r}
tail(AirlineDelays)
```

## Dataset 2

<https://github.com/YBI-Foundation/Dataset/blob/main/Airline%20Delay.csv>

```{r}
AirDelays <- read.csv("Airline Delay.csv")
head(AirDelays)
```

```{r}
dim(AirDelays)
```

```{r}
tail(AirDelays)
```

This dataset contains detailed information about flight arrivals and
delays for U.S. airports, specifically for the months of December 2019
and December 2020. Here are the key components of the dataset:

-   **Dimensions:** The dataset consists of 3,351 rows and 21 columns,
    capturing data for multiple airlines and airports.

-   **Columns:**

    -   **year, month:** The year and month of the recorded flights.

    -   **carrier, carrier_name:** The airline's code and full name.

    -   **airport, airport_name:** The airport code and full name.

    -   **arr_flights:** The total number of arriving flights.

    -   **arr_del15:** The number of flights delayed by over 15 minutes.

    -   **carrier_ct, weather_ct, nas_ct, security_ct,
        late_aircraft_ct:** Counts of delays attributed to various
        causes.

    -   **arr_cancelled:** Number of canceled flights.

    -   **arr_diverted:** Number of diverted flights.

    -   **arr_delay:** Total delay time in minutes.

    -   **carrier_delay, weather_delay, nas_delay, security_delay,
        late_aircraft_delay:** Breakdown of delays by specific causes.

-   **Insights:**

    -   The dataset allows for analysis of flight performance by
        different carriers across various airports.

    -   It can be used to identify trends in delays, cancellations, and
        the reasons behind them, providing insights into factors
        affecting the aviation industry during these months.
