---
title: "Data Exploration"
author: "Heather Anderson"
format:
  html:
    link-external-icon: true
    link-external-newwindow: true
course: Capstone Projects in Data Science
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
theme: cyborg
---

## Dataset 1

<https://www.kaggle.com/datasets/sriharshaeedala/airline-delay?resource=download>

```{r}
AirlineDelays <- read.csv("Airline_Delay_Cause.csv")
head(AirlineDelays)
```

```{r}
dim(AirlineDelays)
```

This dataset provides information on flight arrivals and delays at U.S.
airports, sorted by airline carriers. It includes details like the
number of arriving flights, delays over 15 minutes, cancellations, and
diversions. It also shows the reasons for delays, such as airline
issues, weather, airspace factors, security, and late arrivals. You can
use this dataset to analyze how different airlines perform at various
airports and understand what causes delays in the aviation industry.

Variables:

-   year: The year of the data.

-   month: The month of the data.

-   carrier: Carrier code.

-   carrier_name: Carrier name.

-   airport: Airport code.

-   airport_name: Airport name.

-   arr_flights: Number of arriving flights.

-   arr_del15: Number of flights delayed by 15 minutes or more.

-   carrier_ct: Carrier count (delay due to the carrier).

-   weather_ct: Weather count (delay due to weather).

-   nas_ct: NAS (National Airspace System) count (delay due to the NAS).

-   security_ct: Security count (delay due to security).

-   late_aircraft_ct: Late aircraft count (delay due to late aircraft
    arrival).

-   arr_cancelled: Number of flights canceled.

-   arr_diverted: Number of flights diverted.

-   arr_delay: Total arrival delay.

-   carrier_delay: Delay attributed to the carrier.

-   weather_delay: Delay attributed to weather.

-   nas_delay: Delay attributed to the NAS.

-   security_delay: Delay attributed to security.

-   late_aircraft_delay: Delay attributed to late aircraft arrival.

```{r}
tail(AirlineDelays)
```

## Dataset 2

<https://github.com/YBI-Foundation/Dataset/blob/main/Airline%20Delay.csv>

```{r}
AirDelays <- read.csv("Airline Delay.csv")
head(AirDelays)
```

```{r}
dim(AirDelays)
```

```{r}
tail(AirDelays)
```

All of the data from Dataset 2 appear to be from December 2019 and
December 2020.
