---
title: "Bayesian Linear Regression"
subtitle: "Heather Anderson & Sara Parrish"
author: "Bayesian Bandits"
execute:
  echo: true
  warning: false
  message: false
  error: false
format: 
  revealjs:
    theme: dark
    embed-resources: true
    slide-number: false
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
editor: source
pdf-separate-fragments: true
fig-align: center
---


::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

In kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]\*

## Literature Review: Introduction to Bayesian Linear Regression

-   BLR definition and its significance in statistical modeling

-   Bayesian methods vs. Frequentist approaches

-   Role of prior knowledge in shaping predictions

## Literature Review: Understanding the Bayesian Framework

-   Explanation of Bayes’ theorem and its application in updating beliefs

-   Posterior distribution vs. point estimates

-   How this leads to better uncertainty quantification

## Literature Review: Tools and Techniques in Bayesian Analysis

-   Introduction to the Bayesian Analysis Toolkit (BAT) and its features

-   Markov Chain Monte Carlo (MCMC) for sampling and model comparison

-   Visualization tools (trace plots, HMC diagnostics) for model assessment

## Literature Review: Advantages of Bayesian Methods

-   Findings from Zyphur and Oswald on reliability with small samples

-   Advancements in computing and their impact on Bayesian relevance

-   Efficiency of variational inference as an alternative to MCMC

## Literature Review: Conclusion and Future Directions

-   Key strengths of BLR: flexibility, uncertainty management

-   Encouragement of exploration of MCMC and foundational Bayesian principles

-   Potential for future applications in complex data analysis


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
